---
title: "Explore Factors that determine the red wine quality"
author: "Fienny Angelina"
date: "28 Januari 2019"
output: 
  html_document:
    # theme of html document
    # theme of code highlight                                 
    # table of contents
    theme       : journal            # "default", "cerulean", "journal",
                                    # "flatly", "readable", "spacelab",
                                    # "united", "cosmo", "lumen", "paper", 
                                    # "sandstone", "simplex", "yeti"
    highlight   : tango          # "default", "tango", "pygments",
                                    # "kate",  "monochrome", "espresso",
                                    # "zenburn", "haddock", "textmate"
    toc         : true              # get table of content
    toc_depth   : 2
---

========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

# knitr: Suppress code/messages/warnings 
#        Set default plot options and center them
knitr::opts_chunk$set(fig.width=9,fig.height=5,fig.path='Figs/',
                      fig.align='center',tidy=TRUE,
                      echo=FALSE,warning=FALSE,message=FALSE)

# the library used in this file
library(ggplot2)
library(GGally)
library(gridExtra)
library(memisc)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wineData <- read.csv('wineQualityReds.csv')

dim(wineData)
wineData$quality <- as.factor(wineData$quality)
wineData <- subset(wineData, select = -c(X))
str(wineData)

summary(wineData)
```

Our dataset consists of 13 variables, with 1599 observations. There are no null
values in all columns and we have changed the quality to be factor instead of
integer and removed X column from the dataset. In the end we are left with 12 
variables.

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
# Histogram - Quality
ggplot(aes(x=quality), data=wineData) + geom_histogram(stat="count") + 
  scale_y_continuous(breaks=seq(0, 700, 50))
```


From the above, we can see that the quality of red wine data we have spanned 
from 3 to 8. We can also see that most of the red wine data we have have quality
of 5 or 6. There are very small number of data where the quality is 8 or 3.
We also try to have more detailed scale in the y direction to see how imbalance
the quality data is. From there we can observed that there are around 40 wine 
data with the quality score 4.

```{r echo=FALSE}
# Histogram - pH
ggplot(aes(x=pH), data=wineData) + geom_histogram()
# Histogram - pH
ggplot(aes(x=pH), data=wineData) + geom_histogram(binwidth=0.01)

```

From the above we can see that most of the red wine have pH of 3.2 to 3.4. Also,
the pH plot above looks like a normal distribution. Moreover, there seems to be
outliers for the pH level, those that are above 4 and below 2.75.

```{r echo=FALSE}
# Histogram - alcohol
ggplot(aes(x=alcohol), data=wineData) + geom_histogram()
# Histogram - density
ggplot(aes(x=density), data=wineData) + geom_histogram()

```

From the alcohol data we can see that most wine has alcohol percentage of 9%. We
can also observed that most of the wine has density of 0.995 to 1 and that the
density distribution looks like a normal distribution.

```{r echo=FALSE}
# Histogram - citric acid
p1 <- ggplot(aes(x=citric.acid), data=wineData) + geom_histogram()
# Histogram - citric acid with smaller binwidth
p2 <- ggplot(aes(x=citric.acid), data=wineData) + geom_histogram(binwidth=0.01)

library(gridExtra)

grid.arrange(p1, p2, ncol=1)
```

It seems that most of the data has citric acid of near 0. However, we also see
quite a considerable amount around 0.5. The graph above looks right skewed, we 
may need to use log transformation later.

```{r echo=FALSE}
# Histogram - fixed acidity
p1 <- ggplot(aes(x=fixed.acidity), data=wineData) + geom_histogram(binwidth=0.1)
# Histogram - volatile acidity
p2<- ggplot(aes(x=volatile.acidity), data=wineData) + 
  geom_histogram(binwidth=0.01)

grid.arrange(p1, p2, ncol=1)

```

The most common fixed acidity value lies between 6 - 8, while the most common
volatile acidity value lies between 0.4 to 0.8.

```{r echo=FALSE}
# Histogram - residual sugar
ggplot(aes(x=residual.sugar), data=wineData) + geom_histogram(binwidth=0.1)
# Histogram - sulphates
ggplot(aes(x=sulphates), data=wineData) + geom_histogram()
```

For residual sugar, the most common value lies between 1 - 3 g/dm^3. 
For sulphates, the most common value lies between 0.5 to 1 g/dm^3.

We can also see that there are several data with residual sugar measurement 
more than 8 and total sulfur dioxide of more than 200. These could potentially
be an outlier.

```{r echo=FALSE}
# Histogram - free sulfur dioxide
ggplot(aes(x=free.sulfur.dioxide), data=wineData) + geom_histogram(binwidth=1)
summary(wineData$free.sulfur.dioxide)
# Histogram - total sulfur dioxide
ggplot(aes(x=total.sulfur.dioxide), data=wineData) + geom_histogram(binwidth=2)
summary(wineData$total.sulfur.dioxide)
# Histogram - sulful dioxide percentage
wineData$free.sulfur.dioxide.percentage = wineData$free.sulfur.dioxide / 
  wineData$total.sulfur.dioxide
ggplot(aes(x=free.sulfur.dioxide.percentage), data=wineData) + geom_histogram()


```

For free sulfur dioxide, the most commmon value lies between 0 to 15 mg/dm^3.
For total sulfur dioxide, the most common value lies between 0 to 50 mg/dm^3.
After that, we also create the variable free sulful dioxide percentage which is
the free sulfur dioxide divided by the total sulfur dioxide. All the value are 
between 0 to 1.

```{r echo=FALSE}
# Histogram - chlorides
ggplot(aes(x=chlorides), data=wineData) + geom_histogram(binwidth=0.01)
summary(wineData$chlorides)
```

Most wine has chlorides of around 0.1 g/dm^3.

# Univariate Analysis

### What is the structure of your dataset?
Our dataset consists of 13 variables, with 1599 observations. All of the
variables are number.
- Most of the red wine has quality of 5 - 6.
- Most of the variables, like chlorides, sulphates, volatile acidity seems to 
have normal distribution.
- The citric acid variable seems to be skewed to the right. We may need to use 
log transformation later for it.

### What is/are the main feature(s) of interest in your dataset?
The main features in the data set are alcohol and quality. I’d like to determine 
which features are best for predicting the quality of a red wine. I suspect 
alcohol and some combination of the other variables can be used to build a
predictive model to determine the quality of red wine.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
Some other variables that indicate the taste of the wine, such as acidity, sugar
, and density are likely to affect the quality of the red wine. I think density 
and alcohol contribute most to the quality after researching information on wine
quality.

### Did you create any new variables from existing variables in the dataset?
I create free sulfur dioxide percentage that measure the percentage of free 
sulfur dioxide among all sulfur dioxide in the dataset.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
I checked each column to see if there is any null value, but apparently there
is not. I also change the quality to be a factor instead of number.

# Bivariate Plots Section

```{r echo=FALSE, fig.width=10, fig.height=8}
# Plot - correlation
pair <- ggpairs(wineData) +
  theme(text = element_text(size=10), 
     axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size=7))
pair

```

From the plot above, we can see that alcohol, sulphates, and volatile acidity 
seem to have weak correlation with quality. We can also clarify our previous 
findings that some of the variables have right skewed distribution. 

However, it is interesting to see that there are strong correlation between 
other 2 variables, such as fixed 
acidity vs citric acid, volatile acidity vs citric acid, fixed acidity vs 
density, fixed acidity vs pH. Even though we are interested in only what factors
correlate with the quality, it is also important to take note which pair of 
variables seems to be correlated, especially if later we would like to do a 
linear regression.

I want to look closer at the scatter plots involving quality and some other
variable like alcohol, volatile acidity and sulphates.

Since it is typically not a good idea to use scatter plot for discrete data,
we need to use jitter to avoid overplotting of the plot due to points in 
visualizatoins plotting on top of each other.

```{r echo=FALSE}
# boxplot + scatterplot - alcohol ~ quality
ggplot(aes(x=quality, y=alcohol), data=wineData) + geom_jitter(alpha=0.5) + 
  geom_boxplot(alpha=0.25, color="darkblue") + stat_summary(fun.y=mean, geom='point', shape=4, color="red")
qualAlcohol <- lm(as.numeric(quality) ~ alcohol, data=wineData)
summary(qualAlcohol)
by(wineData$alcohol, wineData$quality, summary)
```

After adding jitter to the plot, we can see that on most cases, the red wine
alcohol quantity seems to correlate to the quality of the wine. The R-squared
value shows that alcohol explain about 22.63 percent of the quality of the wine.

Let's try the same thing for the volatile acidity

```{r echo=FALSE}
# boxplot + scatterplot - volatile acidity ~ quality
ggplot(aes(x = quality, y = volatile.acidity),
      data = wineData) +geom_jitter(alpha=0.3) + geom_boxplot(alpha=0.25, 
      color="darkblue") + stat_summary(fun.y=mean, geom='point', shape=4, 
      color="red")

qualVol <- lm(as.numeric(quality) ~ volatile.acidity, data=wineData)
by(wineData$volatile.acidity, wineData$quality, summary)
```

After setting the transparency, there seems to be a weak negative linear 
correlation between volatile acidity and quality. It is becoming more clear when
we plot it as a boxplot. Nearly 50% of the wine with lowest quality seems to
have higher volatile acidity than other red wine with greater quality. 
In fact, volatile acidity explains 15% of the quality of red wine. We can see 
that as quality increase, the mean and median of red wine belonging to that
quality decrease. From the boxplot, since most of the value range overlap each
other for different quality, volatile acidity may only be used to differentiate 
wine of quality 3 and wine of quality 7 to 8.

```{r echo=FALSE}
# boxplot + scatterplot - sulphates ~ quality
ggplot(aes(x=quality, y=sulphates),      
      data = wineData) +geom_jitter(alpha=0.3) + geom_boxplot(alpha=0.25, 
      color="darkblue") + stat_summary(fun.y=mean, geom='point', shape=4, 
      color="red")

qualVol <- lm(as.numeric(quality) ~ sulphates, data=wineData)
summary(qualVol)
```

We can also see correlation of sulphates vs quality and that the median of 
the sulphates content in the wine increase as quality increase. However, there
seems to be considerable amount of outlier when the quality is 5 and 6. It may
be because of the fact that we have a lot more data for wine of quality 5 and 6
or it could be an indication that sulphates might not be a strong indicator 
of quality. In fact, sulphates explains only about 6% of quality.

```{r echo=FALSE}
# boxplot + scatterplot - free sulfur dioxide ~ quality
ggplot(aes(x=quality, y=free.sulfur.dioxide),
      data = wineData) +geom_jitter(alpha=0.3) + geom_boxplot(alpha=0.25, 
      color="darkblue") + stat_summary(fun.y=mean, geom='point', shape=4, 
      color="red")
```

From the boxplot and scatterplot we draw above, it does not seem that free
sulfur dioxide has linear correlation with quality.

Let's try to draw the correlation between fixed acidity and density

```{r echo=FALSE}
# Scatterplot - density ~ fixed acidity
ggplot(aes(x=fixed.acidity, y=density), data=wineData) + geom_point() + 
  geom_smooth(method="lm", color="blue")

qualVol <- lm(density ~ fixed.acidity, data=wineData)
summary(qualVol)
```

It seems like there is weak linear relation between fixed acidity and density
based on the graph. In fact, fixed acidity explains 45% of density.

We know in fact, at free SO2 concentrations over 50 ppm, SO2 becomes evident in 
the nose and taste of wine, while at low concentration, it is undetectable.
Let's find out how many sample we have had high concentration of SO2 and its 
summary for each of quality value.

```{r echo=FALSE}
# See the number of sample in which the free sulfur dioxide is greater than 50 
# ppm
summary(wineData$free.sulfur.dioxide > 50)
# See the number of sample in which the free sulfur dioxide is greater than 50 
# ppm for each quality factor.
by(wineData$free.sulfur.dioxide > 50, wineData$quality, summary)
```

It seems that in our sample data, only red wine of quality 5 - 7 have sample of
wine with high concentration of S02. However, the number of cases itself is
small compared to the sample size we have, thus it may not be wise to assume that
in general, when there is high concentration of SO2, the red wine is of quality
5 - 7.

Let's look at the correlation of fixed acidity vs pH

```{r echo=FALSE}
# Scatterplot - pH ~ fixed acidity
ggplot(aes(x=fixed.acidity, y=pH), data=wineData) + geom_point() + 
  geom_smooth(method="lm", color="blue")

qualVol <- lm(pH ~ fixed.acidity, data=wineData)
summary(qualVol)
```

It seems that there is weak negative correlation between pH and fixed acidity.
It makes sense as when the wine contains more acids, it has less pH.

Let's see the correlation between volatile acidity and alcohol to determine if
we can use both variable to predict the quality of the wine.

```{r echo=FALSE}
# Scatterplot - alcohol ~ volatile acidity
ggplot(aes(x=volatile.acidity, y=alcohol), data=wineData) + geom_point() + 
  geom_smooth(method="lm", color="blue")

qualVol <- lm(volatile.acidity ~ alcohol, data=wineData)
summary(qualVol)
```

As we can see from above, the adjusted R-squared between volatile acidity and 
alcohol is small and this indicate that we can use both variable together to 
predict the quality of the wine.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Price correlates weakly with volatile acidity and alcohol.

When alcohol percentage increases, the quality tends to increase. However, as 
quality increase, the alcohol percentage variance also increase.

Based on the R^2 value, alcohol explains about 23% of the variance in quality.
Other feature of interest could be incorporated into the model to explain the 
variance in quality.

Red wine with high volatile acidity concentrate tend to have lower quality. Wine
with quality of 3 all have volatile acidity concentrate higher than 0.7 g/dm^3 
and most of wine with quality greater than 6 have volatile acidity concentrate 
lower than 0.5 g/dm^3. I suppose it is because high acidity lead to unpleasant
taste.

Moreover, wine with higher concentration of sulphates tend to have higher 
quality.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
We can see stronger correlation between fixed acidity vs density 
and between fixed acidity and pH.

### What was the strongest relationship you found?
The relation between fixed acidity and pH. As for the main feature, it will be 
that the quality of wine strongly correlated with alcohol and volatile acidity.
Moreover since alcohol and volatile acidity does not seem to have correlation,
we can use both variable later to predict the quality of the wine.


# Multivariate Plots Section


```{r echo=FALSE, Multivariate_Plots}
# Scatterplot - alcohol ~ volatile acidity
ggplot(aes(x = alcohol, y = volatile.acidity, color=quality),
       data = wineData) + geom_point() + scale_color_brewer(type="div")

# Scatterplot + lineplot of median - alcohol ~ volatile acidity
ggplot(aes(x = alcohol, y = volatile.acidity, color=quality), data = wineData) +
  geom_point() + 
  geom_line(stat = 'summary', fun.y = median) + scale_color_brewer(type="div")


# Scatterplot + lineplot of median - alcohol ~ volatile acidity with rounding
ggplot(aes(x = round(alcohol), y = volatile.acidity, color=quality), data = 
  wineData)+ geom_line(stat = 'summary', fun.y = median) +
  scale_color_brewer(type="div")

```

In the plot above we plot the graph of median of volatile acidity vs alcohol
for each quality value. In the graph above we can see that as quality decrease 
to 1, the volatile acidity tend to increase for some value of alcohol. However, 
it does not hold for all alcohol percentage, especially when the alcohol 
percentage is above 12. From the above graph, we can also see that each line
looks random which show that there is no linear relation between alcohol and
volatile acidity.

```{r echo=FALSE}
# Scatterplot - pH ~ fixed acidity, colored by quality
ggplot(aes(x = fixed.acidity, y = pH, color=quality),
       data = wineData) + geom_point() + scale_color_brewer(type="div")
```

In the plot above, it seems that even though there is correlation between fixed 
acidity and pH, there is no clear separation of the color of point in the plot.
It shows that both variable combined does not correlate to quality

```{r echo=FALSE}
# Scatterplot - density ~ alcohol, colored by quality
ggplot(aes(x = alcohol, y = density, color=quality),
       data = wineData) + geom_point() + scale_color_brewer(type="div")
```

This plot looks a bit better than the previous one as in there, we can see a bit
of grouping in the location of each color.

```{r echo=FALSE}
m1 <- lm(as.numeric(quality) ~ alcohol, data=wineData)
m2 <- update(m1, ~ . + volatile.acidity)
m3 <- update(m2, ~ . + sulphates)
m4 <- update(m3, ~ . + citric.acid)
m5 <- update(m4, ~ . + density)

mtable(m1, m2, m3, m4, m5)

```
# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Alcohol and volatile acidity seems to strengthen each other. Moreover, density
seems to also strengthen alcohol.

### Were there any interesting or surprising interactions between features?
The interaction between alcohol and density seems to able to contribute to the
quality of the wine. However, it is not strong enough as some different quality
value still override each other in the scatter plot we draw.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

Yes, I created 5 model to predict the quality of the wine. At first, I only use 
alcohol, in which I obtain 0.226 as the adjusted R-squared. After that, I added
volatile acidity, after which the adjusted R-squared increase a lot while making
all the variables as still important. After that, I added sulphates, which also
increase the R-squared, while keeping all variable important. After that, I 
added citric acid and density, after which the R-squared did not improved a lot
and the variables added is marked as unimportant. 

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
# Histogram - quality
ggplot(aes(x=quality), data=wineData) + geom_histogram(stat="count", fill="darkblue") + 
  scale_y_continuous(breaks=seq(0, 700, 50)) + labs(y="Number of Observations",
  x="Wine Quality", title="Distribution of red wine quality") +
  scale_color_brewer()
```

### Description One

From the plot above, we can see that our data of wine quality is imbalanced. 
There are a lot more ordinary wine (those with quality of 5 - 6) than the good
ones ( quality of 7- 8 ) or the bad ones (3 - 4). It is probably because the
ordinary wine is the most popular one as it is not as expensive as the good ones
but is still tasty.

### Plot Two
```{r echo=FALSE, Plot_Two}

# Boxplot - alcohol differentiated by quality
ggplot(aes(x = quality, y = alcohol), data = wineData) + 
  geom_boxplot(fill="darkgreen") + stat_summary(fun.y=mean, geom='point', 
                                                shape=4) +
  labs(title="Alcohol percentage for each quality of wine",
       x="Quality",
       y="Alcohol (% by Volume)")

```

### Description Two
In this plot, there seems to be a tendency for alcohol percentage to increase
as the quality of wine increase. However, there is some anomality when the 
quality is 5, in which the alcohol percentage average is lower than when 
the quality is 4. Moreover, there are also some outlier spotted when the 
quality is 5.

### Plot Three
```{r echo=FALSE, Plot_Three}
# Line plot - volatile acidity ~ alcohol, differentiated by quality
ggplot(aes(x = round(alcohol), y = volatile.acidity, color=quality),
       data = wineData) + geom_line(stat = 'summary', fun.y = median) + 
  scale_color_brewer(type="div") +
  labs(x="Alcohol (% by Volume)", y=expression(Volatile~Acidity~(g/dm^{3})),
       title="Alcohol and Volatile Acidity, by Quality")

```

### Description Three

The plot above shows that it may be possible to predict quality using alcohol
and volatile acidity. It is because when comparing the line for low quality and 
high quality, we can see clear separation, e.g when quality is 3 vs 8. Thus,
we can see that it would be easy to separate low quality wine from the high 
quality ones. However, the relationship may not be a linear. When we want to 
categorise the wine, it may be better to use logistic regression or SVM or 
K-means.

------

# Reflection

The red wine data set contains information on almost 2000 thousand red wines
across 12 variables from around 2009 I started by understanding the individual 
variables in the data set, and then I explored interesting questions and leads
as I continued to make observations on plots. Eventually, I explored the quality
of red wine across many variables and created a linear model to predict the
quality.

There was a clear trend between the alcohol or volatile acidity of wine and its 
quality. I was surprised that pH or free sulfur dioxide did not have a strong 
negative correlation with quality, but these variables are likely to be 
represented by sulphates. I struggled understanding the outliers in the box plot
that usually occured when the quality is 5 or 6, but this became more clear when
I realized that most of the data has quality of 5 to 6. For the linear model,
all red wine were included since information on quality, volatile acidity, 
alcohol, sulphates, citric acidity, and density were available for all row. 
After fitting the linear model without transforming the variables, the model was
able to account for 33.7% of the variance in the dataset.

Some limitation of this model include the fact that we use linear regression to 
fit a factor variable. In this case, I think it may be better to separate the 
quality into 2 factor, good or bad and fit a logistic regression / SVM / K-means
model into the data. It is because it is easier to separate good and bad wine 
first before we mark the quality of the wine.